---
title: "Pakistan: From Stabilization to Transformation"
subtitle: "A Growth Diagnostics Brief"
author: "Growth Lab, Harvard Kennedy School"
date: "April 2026"
output:
    html_document: default
bibliography: references.bib
---

<!--
    Mini-report playground — tests the GL Report Design Framework.
    This file uses content from the Pakistan FM engagement to create
    a realistic 4-page report exercising all the major framework elements:
    headings (H1–H3), body text, figures at different sizes, a table,
    a call-out box, and citations.
-->

```{r setup, message=FALSE, warning=FALSE, results='hide'}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(ggthemes)
library(ggrepel)
library(arrow)
library(showtext)
library(sysfonts)
library(patchwork)
library(scales)

# ---- Growth Lab design tokens ------------------------------------------------

gl <- list(
    text_dark   = "#333333",
    text_muted  = "#7c7c7c",
    border      = "#dcdcdc",
    background  = "#f3f3f3",
    brand_blue  = "#266798",
    palette = c(
        "#266798", "#C64646", "#36B250", "#EAC218", "#D1852A",
        "#52E2DE", "#A42DE2", "#7C6760", "#757777"
    )
)

highlight    <- "#C64646"
highlight_sz <- 1.1

# ---- Fonts -------------------------------------------------------------------

font_add_google("Source Sans 3", "gl_sans")
font_add_google("JetBrains Mono", "gl_mono")
showtext_auto()
showtext_opts(dpi = 300)

# ---- Theme -------------------------------------------------------------------

theme_gl <- function(base_size = 12) {
    theme_few(base_size = base_size) %+replace%
    theme(
        text          = element_text(family = "gl_sans", color = gl$text_dark),
        plot.title    = element_text(family = "gl_sans", face = "plain",
                                     size = rel(1.35), hjust = 0,
                                     margin = margin(b = 4)),
        plot.subtitle = element_text(family = "gl_sans", color = gl$text_muted,
                                     size = rel(1.0), hjust = 0,
                                     margin = margin(b = 10)),
        plot.caption  = element_text(family = "gl_mono", color = gl$text_muted,
                                     size = rel(0.75), hjust = 1,
                                     margin = margin(t = 8)),
        axis.title    = element_text(family = "gl_mono", color = gl$text_muted,
                                     size = rel(0.85)),
        axis.title.x  = element_text(margin = margin(t = 6)),
        axis.title.y  = element_text(margin = margin(r = 6), angle = 90),
        axis.text     = element_text(family = "gl_sans", color = gl$text_dark,
                                     size = rel(0.85)),
        legend.title  = element_text(family = "gl_mono", color = gl$text_muted,
                                     size = rel(0.8)),
        legend.text   = element_text(family = "gl_sans", size = rel(0.85)),
        strip.text    = element_text(family = "gl_mono", color = gl$text_dark,
                                     size = rel(0.85))
    )
}

theme_set(theme_gl())

options(
    ggplot2.discrete.colour = gl$palette,
    ggplot2.discrete.fill   = gl$palette
)

# ---- Figure sizes (from recipes/report.md) ----------------------------------------

gl_fig <- list(
    full        = list(w = 6.5,   h = 4.0),
    full_tall   = list(w = 6.5,   h = 6.0),
    full_square = list(w = 6.5,   h = 6.5),
    major       = list(w = 4.278, h = 4.0),
    half        = list(w = 3.167, h = 3.0),
    half_tall   = list(w = 3.167, h = 5.0)
)

save_fig <- function(name, filename, plot = last_plot(), dpi = 300) {
    sz <- gl_fig[[name]]
    ggsave(file.path("imgs", filename), plot = plot,
           width = sz$w, height = sz$h, dpi = dpi)
}

# ---- Data --------------------------------------------------------------------

macro_df <- read_parquet("~/dev/shared-data/growth-lab/glmacro_master_alldata.parquet")

main_country     <- "PAK"
main_comparators <- c("BGD", "EGY", "IND", "IDN", "LKA", "MAR", "VNM")

```

# Introduction

Pakistan stands at a critical juncture. After decades of boom-bust cycles
driven by recurring balance-of-payments crises, the country has entered a new
stabilization phase under its latest IMF program. But stabilization alone is
not a growth strategy. Without a deliberate shift toward export-led growth and
productive diversification, Pakistan risks returning to the same pattern that
has kept its per capita income stagnant relative to peers.

This brief presents a diagnostic overview of Pakistan's macroeconomic
trajectory, drawing on data from the IMF, World Bank, and the Atlas of
Economic Complexity. It accompanies the Growth Lab's proposed 18-month
engagement with the Ministry of Finance.

> **Key finding:** Pakistan's growth above 2% per capita has historically been
> associated with current account deterioration — a structural constraint that
> can only be relaxed through export diversification.

# The Macro-Fiscal Picture

## Foreign reserves: a recurring vulnerability

Pakistan's foreign exchange reserves have repeatedly fallen below the
three-month-of-imports threshold considered the minimum for macroeconomic
stability. Each episode triggers emergency borrowing, IMF programs, and
contractionary adjustment.

```{r fig_reserves}

macro_df |>
    filter(countrycodeiso == main_country, year > 1990) |>
    rename(reserves = wdi_fi_res_totl_mo) |>
    filter(!is.na(reserves)) |>
    ggplot(aes(x = year, y = reserves)) +
    geom_hline(yintercept = 3, linetype = "dashed", color = "grey") +
    geom_line() +
    geom_point() +
    labs(
        title = "Reserves in months of imports",
        subtitle = "Pakistan, 1990–2023. Dashed line = 3-month threshold.",
        x = NULL, y = "Months of imports",
        caption = "Source: World Bank WDI"
    )

save_fig("full", "reserves.png")

```

**Figure 1.** Pakistan's foreign exchange reserves in months of imports, 1990–2023.

*Source: World Bank World Development Indicators.*

## Growth and the current account constraint

The scatter plot below illustrates the core structural problem: years of
higher per capita GDP growth are associated with a worsening current account
balance. This "growth–current account trade-off" is not inevitable — countries
that diversify their exports can grow without generating unsustainable external
deficits.

```{r fig_gdp_comparators}

comparator_set <- c(main_country, main_comparators)

gdppc <- macro_df |>
    filter(countrycodeiso %in% comparator_set,
           year %in% c(2000, 2023),
           !is.na(wdi_ny_gdp_pcap_kd)) |>
    select(countrycodeiso, year, gdppc = wdi_ny_gdp_pcap_kd) |>
    pivot_wider(names_from = year, values_from = gdppc, names_prefix = "y") |>
    mutate(
        growth = (y2023 / y2000) - 1,
        is_pak = countrycodeiso == main_country
    )

gdppc |>
    ggplot(aes(x = y2000, y = growth, label = countrycodeiso)) +
    geom_point() +
    geom_point(data = . %>% filter(is_pak), color = "red", size = 3) +
    geom_text_repel() +
    scale_x_log10(labels = dollar) +
    scale_y_continuous(labels = percent) +
    labs(
        title = "GDP per capita growth, 2000–2023",
        subtitle = "Pakistan vs regional comparators",
        x = "GDP per capita, 2000 (constant USD, log scale)",
        y = "Cumulative growth, 2000–2023",
        caption = "Source: World Bank WDI"
    )

save_fig("full", "gdp_comparators.png")

```

**Figure 2.** GDP per capita growth (2000–2023) against initial income level.
Pakistan highlighted in red.

*Source: World Bank WDI. Authors' calculations.*

## Private credit: a symptom of shallow finance

```{r fig_credit_comparison}

credit <- macro_df |>
    filter(countrycodeiso %in% comparator_set,
           year == 2022,
           !is.na(wdi_fs_ast_prvt_gd_zs)) |>
    select(countrycodeiso, credit = wdi_fs_ast_prvt_gd_zs) |>
    mutate(is_pak = countrycodeiso == main_country) |>
    arrange(desc(credit))

credit |>
    ggplot(aes(x = reorder(countrycodeiso, credit), y = credit)) +
    geom_col() +
    geom_col(data = . %>% filter(is_pak), fill = "red") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(scale = 1)) +
    labs(
        title = "Domestic credit to private sector, 2022",
        subtitle = "Pakistan vs comparators (% of GDP)",
        x = NULL, y = "% of GDP",
        caption = "Source: World Bank WDI"
    )

save_fig("half_tall", "credit_comparison.png")

```

**Figure 3.** Domestic credit to the private sector as a share of GDP,
Pakistan vs comparators.

*Source: World Bank WDI.*

# Toward a Growth Strategy

The Growth Lab's engagement with Pakistan is structured around two pillars:

| Pillar                          | Focus                                    | Timeline    |
|---------------------------------|------------------------------------------|-------------|
| Macro-Fiscal Framework          | Fiscal space, debt sustainability, SBP   | Months 1–8  |
| Export Diversification Strategy  | Product space, complexity, industrial policy | Months 4–18 |

Both pillars are supported by an **Action-Learning Team** model that pairs
Growth Lab researchers with counterparts in the Ministry of Finance, the
Planning Commission, and the State Bank of Pakistan.

> The goal is not to produce a report that sits on a shelf, but to build
> analytical capacity within government institutions that outlasts the
> engagement itself.

# References
